Are populations like a circuit? Comparing isolation by resistance to a new coalescent-based method.

### 2019-04-20 review-responses.tex

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are writing to submit a revised version of our manuscript, 
``Are populations like a circuit? 
Comparing isolation by resistance to a new coalescent-based method.''
(formerly titled 
``Are populations like a circuit? 
The relationship between isolation by distance and isolation by resistance'').

Thanks to the five mostly anonymous reviewers for their many useful comments --
we've addressed all of them, which we think has made the paper more generally understandable
and suitable for this audience.
In particular, we have included analysis of an empirical dataset 
(which was fun, and produced good results),
revised the figures for increased legibility,
and added a substantial amount of additional explanation of the technical points.
We hope that discussion of the empirical results should help make interpretation
of the coalescent-based model more broadly understandable.
And, at the suggestion of reviewer 1,
we have changed the title.
We think this has all improved the paper substantially,
and hope you agree.

We have included detailed responses to all the reviewer's comments,
with page and line numbers that refer to the revised version of the manuscript.
There is also a version of the manuscript showing differences to the previously submitted document.

@@

%%%%%%%%%%%%%%
\reviewersection{AE}

\begin{quote}
We received five reviews for your submitted article. While three are quite positive, two others are more critical. In the balance of comments, all reviewers felt that your article should be published in MER, though significant improvements were requested. Therefore, my decision is to request a revision before accepting your article.

Please respond to all major comments raised by the reviewers. In particular, try to address the following points: 1) Revise figures for clarity and legibility. 2) Include an empirical dataset, as suggested by REV 2. 3) Rewrite for a broader audience of biologists (translating some difficult concepts when possible and as appropriate). 4) Clarify aspects of the statistical methods as indicated by several reviewers.
\end{quote}

Please find our responses below.
We hope you agree we've addressed all four of these points.

@@

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
    The authors present a new Bayesian method to fit movement rates between a
    set of demes to pairwise genetic differentiation, utilizing a MCMC. They first
    state the basic equations, which closely resemble existing theory of stepping
    stone models. They extensively compare their computational method to resistance
    based methods using an own implementation of the latter, both ``under the model''
    as well as using various quite realistic forward in time simulations.
    Resistance based methods are much faster and can handle much more demes than
    the method presented here (no case with more than a 5x5 grid is shown).
    Therefore, as the authors point out, this work will likely not contest the use
    of resistance distances for visualizing spatial genetic patterns (which is
    often done in an exploratory way). But importantly, the method here can infer
    asymmetric and biased rates, unlike almost all existing methods, which presents
    a major advance -- and certainly fills a niche.  
\end{quote}

@@

\begin{quote}
    Despite the computational limitations, I believe that this paper will provide
    researchers a good starting point for future development, by both giving a
    review about difficulties and also a first idea and example for an
    implementation.

    Overall, I think this article makes a very valid contribution for Molecular
    Ecology Resources. The authors put their work into the right context and
    show good knowledge of the relevant literature. The writing is both clear
    and concise. Importantly, they do not shy away from pointing out the limits
    of their method, and the very stringent (often unrealistic) assumptions
    such methods usually make.

@@

\begin{point}{}
    Clearer point out the current computational limitations of your coalescence
time method. It seems it gets tricky with more than about 30 demes? What is the
run-time then, does the whole thing blow up? Give an idea how big is too big!
\end{point}

\reply{
    The runs with 16 demes take around 10 hours with the number of iterations we used
    (noted in the text, \revref).
    Thirty demes is probably a good limit for the current implementation. 
    There are changes that could be made to make it more efficient for larger systems, 
    but at that point the ill conditioning problem would be more severe 
    because there would be more parameters.
    We've added a note about this to the discussion. \llname{ll:computation}
}

\begin{point}{}
    What is the exact scaling? The bottleneck seems to be calculating the
    coalescence times with equation (7). You only say that it ``can be solved
    with standard linear algebra''. Are you doing Gaussian elimination or
    something else worse than $N^2$? Do you use any speed-ups to
    deal with the sparse nature of G for neighbor-only migration? Is it faster
    than getting the coalescence times by iterating time-discretized
    recursions?  It would be helpful for further progress to describe the exact
    scaling and the exact algorithm you use.
\end{point}

\reply{
    In the current implementation, 
    equation (7) is solved using \texttt{solve(A,b)} in R. 
    The actual method varies depending on the specific linear algebra library used, 
    but it is typically Gaussian elimination or LU decomposition, 
    both of which are $\mathcal{O}(N^3)$ 
    (unfortunately the decomposition is different in each step;
    if it were the same it would be $\mathcal{O}(N^2)$).
    For fine discretizations of space, the matrices are sparse,
    but in our testing, using sparse matrix methods didn't help at this scale.
    We've added a point about this last thing to the discussion \revref{}
    but haven't put in more detail (e.g., plots of scaling with $N$)
    because it depends on the details (sparsity of the adjacency matrix),
    and as we discuss above, we think that nonidentifiability is a more serious barrier
    to finer resolutions.
}

@@

\reply{
    This is a good point, and we did experiment with other measures of model fit.
    However, in practice the mean absolute error captured patterns as well as other methods.
    In the interest of not increasing complexity of the paper, we haven't included results
    from the other methods.
}

\begin{point}{}
    You write you use standard methods to check convergence and mixing of the
    MCMC chain \revreffull{1}{6}. Given the degeneracy and the high dimensionality,
    where there any problems? E.g., local optima you were stuck sometimes? Or
    was it always pretty straightforward? Maybe write a few sentences on that!
\end{point}

\reply{
    Convergence could be slow, and the likelihood surface is often ridgy, 
    but we did not see too many problems with getting stuck in completely wrong
    areas of state space.
    This the reason we introduced the non-standard ``pre-burn-in'' phase.
}

\begin{point}{\revref}
    $F_{ST} \neq$ Absolute divergence!!!  $F_{ST}$ is more a
    co-variance measure, and for instance heavily influence by sampling a batch
    of related individuals (whereas the mean of absolute divergence will be
    unchanged).
\end{point}

\reply{
    We can't agree more!!!!
    We've added a note about this, although not a lot of detail,
    because it seems outside the scope of this paper \revref{}.
    And, we thought we should mention $F_{ST}$ because it is commonly used
    in these situations.
}

@@

\reply{
    Thank you for catching this.
    You are correct that in addition to $ZC$ being skew-symmetric,
    the rows of $Z$ must also sum to zero, which we now say. {\revref}
    Further exploration of the constraints would be interesting,
    but we think is beyond the scope of this paper.
}

\begin{point}{Eq. 4:}
    $H$ may not be always invertible ($H$ symmetric and positive off-diagonal
    and 0 on the diagonal is NOT enough). So there might be some pathological
    cases where multiple G link to one $H$, maybe even in some simple toy
    examples.
\end{point}

\reply{
    Good point; we had not justified this.
    Under the standing assumption that the chain is irreducible, 
    $H$ is always invertible.
    We now prove this in the Appendix \llname{rr:H_invertible},
    and have made reference to this fact. \revref
}

\begin{point}{}
    Dimensionality arguments (\revref{} and onward) can be a bit misleading when you
    already know the equations have a non-trivial kernel. Setting all
    non-neighboring elements of $G$ to 0 might not get rid of parts of the
    kernel. So maybe flag it even more as ``be cautious'' as you already have.
\end{point}

@@

\begin{point}{Fig. 7:}
    With EEMS one can see very suspiciously the 4x4 sampling structure - there
    are red dots in almost all 3x3 centers of the sampling squares. Since EEMS
    is quite fast, could you increase the resolution of sampling?
\end{point}

\reply{
    More samples could be added for the EEMS runs, 
    but doing so would have the drawback 
    that the two methods would no longer be using the same data for inference. 
    We've been told that similar patterns show up when EEMS tries to approximate
    anisotropic migration,
    so it may not be the sampling pattern.
    We've added a note about this. {\revref}
    Also, note that these samples are more spread out that people often have in practice:
    they do occur in the middle three-quarters of each grid square,
    but they aren't all located at the center.
}

\begin{point}{}
    You could experiment with visualizations a bit. EEMS got quite popular as
    exploratory visualization tool because it produces visually pleasing plots.
    Do lines with two arrowheads of different sizes near the middle, maybe even
    colored (blue to red) work?
\end{point}

\reply{
    True! We've tweaked the way the arrows are displayed,
    and think they look better now.
}

@@

\reviewersection{2}

\begin{quote}
    This work revisits McRae’s seminal isolation-by-resistance (IBR) model through
    the lens of coalescent theory. The original idea of IBR was to compare patterns of
    genetic distance observed on a landscape with the so-called ``resistance distance''.
    The resistance distance between states $i$ and $j$ in a Markov chain is defined to be
    the ``commute time'': the expected amount of time it takes for the chain to go
    from $i$ to $j$ and back, integrated over all possible paths of the chain. Resistance
    distance comes from electrical circuit theory and is not a model of genetic
    variation across a landscape. Also, it is inherently symmetric in $i$ and $j$, and
    cannot accommodate asymmetric migration between demes. A more biologically
    realistic model is the structured coalescent. Hence, if data are generated under
    the something resembling the coalescent, then there is scope for improving IBR
    by reducing model misspecification.

    This paper implements such a method and conducts comparative analyses on a
    number of simulated data sets. It seems that in general, this method and the
    resistance-based approaches (as exemplified by EEMS) produce somewhat
    comparable results so long as migration rates are symmetric. The largest
    differences emerge when gene flow is asymmetric, because this fundamentally
    violates the resistance distance model used by EEMS. Here, the authors show
    convincingly that inferring migration rates using the resistance distance
    approximation can be wildly inaccurate compared to the same estimates
    obtained using coalescent methods. To the extent that gene flow is more
    likely to be asymmetric in many natural populations, this seems like an
    important finding.

@@

    My overall impression of this work is fairly positive. The paper is very clearly
    written, and the idea of replacing IBR with a method that is more faithful to
    the underlying biology is appealing. The methods section (as well as appendices)
    feature a very nice and readable summary of the mathematical innovations of
    this work, as well as some interesting discussion of issues of identifiability and
    power. Similarly, discussion section in very well done and really strengthens
    the manuscript. It includes a thorough survey of the related work, as well as a
    forthright discussion of advantages and disadvantages of the proposed method,
    and it raises a number of interesting questions and directions for future research.
    Overall, the paper was a pleasure to read.
\end{quote}

Thanks very much for the kind words.
One minor comment: resistance distances are not strictly based in a population genetics model,
but we think that in practice it is thought of as (an approximation to) such a model
-- for instance, it is common to see the explanation that it
``integrates over all possible paths that the organisms might take''.
We think \citet{mcrae2006isolation} saw resistance distance as an approximation
to the more-accurate coalescence models; we're looking at the effects of this approximation.
We hope this point comes through in the paper.

@@

\begin{quote}
    I find the main deficiency to be the lack of any analysis of real data. All
    experiments are carried out on simulated data only. This is a puzzling omission.
    The authors went to the trouble of creating an R package (which is good), so
    clearly they intend for the method to be used in practice. Personally, I would
    hesitate to publish the results obtained from a method which had not previously
    been verified to produce reasonable answers in real data. This really feels like
    something the authors ought to have included. It is not necessary that the data
    be novel -- reanalyzing (say) the African elephant dataset in the EEMS paper
    would provide for an interesting comparison (even if it cannot be done at an
    equivalent spatial resolution) and would also go long way towards reassuring
    practitioners.
\end{quote}

We've now added analysis of a real dataset, from \citet{geraldes2014landscape},
and are quite pleased with the results!
This has also given us a chance to spend more time talking through how to interpret the results.
% (We didn't get to say this in the paper,
% but initially were worried by the southward bias in gene flow for the inland poplar species,
% since we would expect the opposite if poplars had expanded to the north following the glaciers;
% but then we learned that this species likely expanded from Alaska, rather than from the south.)

@@

\begin{point}{}
    The discussion of ill-conditioning is welcome. From experience, it often
    feels like coalescent-based methods have little power to accurately
    estimate migration unless the rates are large. It might be interesting to
    shift emphasis from estimation to the slightly easier problem of testing:
    how good are these methods at distinguishing between a given null and an
    alternative hypothesis (say, barrier vs.\ no barrier) for the rate matrix $G$?
    I think the gain of using a coalescent model might be more apparent in this case.
\end{point}

\reply{
    This is a good point; we've added it to the discussion
    (and credited you with it). \revref
}

\begin{point}{}
    This method is computationally limited by equation (7), which has a
    complexity of $O(d^6)$ for the product Markov chain on d demes mentioned a
    few lines below. At least, that is the cost for solving (7) exactly -- but
    given the noisiness of the matrix C, there may be little point in doing
    that anyways. I wonder if the authors have thought at all about obtaining
    an approximate solution to (7), where the error in approximation is chosen
    to in some way to balance out the statistical error inherent in C / D. It
    may also be possible to leverage the fact that the solution G is sparse
    here.  (This is more of a suggestion for future research rather than a
    suggestion for revision.)
\end{point}

\reply{
    Thanks for the suggestions.
    Solving (7) directly for $G$ and $\gamma$ with a noisy $C$ is extremely difficult
    even in cases small enough to just use least squares; 
    it typically resulted in some negative values unless there is very little noise in $C$.
    In order to solve for $C$ from $G$ and $\gamma$, 
    we experimented with the Jacobi method. 
    While it seemed to converge fairly reliably, 
    at the grid sizes where it was computationally advantageous to use it over solving exactly, 
    it would still take a very long time for the MCMC to converge
    due to the large number of parameters.
    Something along these lines might be useful for finer geographical resolutions, though.
    % Developing a way to reduce the number of parameters in a way similar to EEMS
    % while still allowing asymmetric migration is a good direction for future implementations.
}

@@

\begin{point}{\revref}
    ``if fecundity is variable'' -- on what scale?  Among populations?  Among
    individuals?  Among ages within the same individual? 
\end{point}

\reply{
    We mean among individuals (and have added this). \revref
}

\begin{point}{\revref}
    ``It is not required in this approach to have samples from every spatial
    location \ldots''  Required for what?  Are they saying that unsampled but real
    (ghost) populations have no effect on precision or bias?
\end{point}

\reply{
    We're saying that it is possible to \emph{run} the method
    to have samples in each location --
    so it is easy, in principle, to include ``ghost'' populations.
    Of course, unsampled populations will affect accuracy of the result,
    and we've added a note to clarify. \revref
    % In practice, the more locations with samples you have, the better the result will be.
    % We would not typically recommend doing inference that includes unsampled demes
    % if it could be avoided.
    % However, in the theoretical case of a population fitting the model perfectly
    % and knowing the other genetic distance values perfectly, 
    % it is possible to solve exactly for paramters relating to the unsampled populations
    % as long as there are as many unique equations as unknowns.
    % In a real situation, if data is very good for the sampled locations, 
    % inference should be reasonable for the unsampled ones.
}

@@

\begin{point}{}
    Many of the figures are too small to read easily without a magnifying glass.
    Figures 3-5 are large enough but they include miniscule fonts that make it
    impossible to tell exactly what is plotted.  Green and black do not provide
    enough contrast.
\end{point}

\reply{
    We sympathize, and have increased the sizes of most of the labels.
    However, the x-axis labels of figures like Figure~\ref{fig:5x3b_post_coalvcom}
    cannot be made larger without overlapping, and we'd like to keep them.
    Another approach would be to simply plot true versus observed values,
    which is equivalent to the current plot without the x-axis labels.
    We prefer to leave these labels, even if these are only really legible
    in the digital version (after zooming in).
}

@@

\begin{point}{\revref}
    the ``wrong answer'' compared to what?
\end{point}

\reply{
    The truth. We've changed the phrasing. \revref
}

\begin{point}{\revref}
    What is EEMS?
\end{point}

\reply{
    A lengthy acronym (that we've added, as well as MAPS, later). \revref
}

@@

\begin{point}{}
    ``suppose that the spatial arrangement of populations is given by a graph
    \ldots''  ``Graph'' is a very general term and not clear exactly what is meant
    here.  If the word has a more specific meaning in this context, please
    define it.
\end{point}

\reply{
    Done. \revref
}

\begin{point}{Fig 6 caption:}
    the barrier is ``sufficiently thick'' to prohibit migration.  ``Thick''
    implies distance, but is that a requirement?  Couldn’t the barrier be
    higher or deeper or less penetrable for a different reason than thickness?
\end{point}

\reply{
    It could, but this is the way we implemented it in the simulation.
}

@@

    The types of questions and the type of demographic scenarios under which
    resistance vs. coalescense based approaches may be an improvement, or
    combined in the same study should be made more clearly. I suggest a
    summarising table or flow chart to this effect, that addresses time scales
    also.
\end{quote}

We've added more discussion along these lines
particularly in {\llname{ll:history}} 
and in relation to analysis of the poplar data.
We've got quite a few figures already (and have added one for the poplars),
so haven't tried to add a flow chart.

\begin{point}{Line 7 of abstract and generally:}
    This statement is conditional on the characteristics of the data you have, and
    your study aim. If the study aim is to  ask how landscape features affect
    genetic distance, then the coalescent time is not going to explicitly test
    this, or? It needs to be made clear how different study objectives will
    benefit in landscape genetics and under what circumstances a `more
    accurate' predictor of genetic distance can be gained. It seems appropriate
    to mention that this approach would greatly complement existing landscape
    genetic analysis approaches to see the relative effects of resistance and
    coalescent time. Implications for management decisions based on resistance
    models could be mentioned, as resistance models are intended to inform
    conservation decisions in some cases.
\end{point}

### 2019-04-20 isolation_by_coalescence.tex
updated figs
final read-through

The \emph{commute time} of a Markov chain between locations $i$ and $j$
is the expected time for a particle that moves according to the rules of the chain, 
started at $i$, to first reach $j$ and then return.
We will denote this by $R_{ij}$, to emphasize that it is also the effective resistance
between $i$ and $j$.
The \emph{coalescence time} is the expected time for two particles started at $i$ and $j$
to \emph{coalesce},
which happens at a given rate whenever they are in the same location.
We will denote this by $C_{ij}$.
Why should these be related to genetic distance? 
First, we explain this connection,
and then discuss their computation,
and finally develop a method that infers movement rates of the Markov chain
(which are reverse-time migration probabilities)
from observations of coalescence time, \revpoint{1}{9}
and apply this to a real dataset.
Although we have tried to highlight the intuitive aspects of this initial technical discussion,
readers who are interested in practical differences
more than the mathematical arguments should be able to skip to ``Test cases and simulations''
without loss of continuity. \llabel{ll:flagging}

@@

For data, we are given genotypes of individuals and their geographic coordinates;
then, we divide these individuals into groups,
and compute the genetic distance between group $i$ and group $j$
as the mean divergence between a pair of genomes, chosen one from each group.
We denote the resulting genetic distance $D_{ij}$;
since divergence is symmetric, $D_{ij} = D_{ji}$.
We compute $D_{ii}$ by sampling without replacement from group $i$,
so that generally $D_{ii} > 0$: the diagonal elements measure local diversity.
As discussed below, we will find commute times so that the mean commute time between
$i$ and $j$ matches $D_{ij}$ -- but the commute time of a location to itself is zero,
so we must account for local diversity.
We do this in the same way as \citet{petkova2016visualizing},
by introducing the \emph{local diversity} parameters $q_i$,
and using the model
\begin{align} \label{eq:commute_approx}
	D_{ij} \approx \comdist_{ij} = R_{ij}/4 + q_{i}/2 + q_{j}/2 .
\end{align}
Concretely, this means finding 
movement rates of a Markov chain whose matrix of commute times ($R$)
yields a good approximation to the matrix of genetic distances,
after adding local diversity parameters ($q$, which are also free parameters). \revpoint{3}{7}
The factor of four is necessary to make commute time work on the same time scale as coalescence time:
the commute time covers the path from $i$ to $j$ twice
(once in both directions),
while the coalescence time covers the distance only once,
and in half the time, because two particles are moving.
% Furthermore, coalescence time includes within-location diversity
% because particles do not coalesce instantly --
% therefore, commute time is most analogous to the first meeting time.
For a conceptual picture of the differences between the coalescence time and commute time,
see Figure \ref{fig:concept_coalcom}.

@@

Now, we explain how to compute the relevant quantities --
mean commute and coalescence times -- from a Markov chain model of lineage movement.
The Markov chain is defined by its \emph{generator matrix}, denoted $G$,
for which $G_{xy}$ gives the ``jump rate'' from $x$ to $y$.
The jump rate $G_{xy}$ is \revpoint{3}{8}
the probability the chain jumps to position $y$ when at position $x$ per unit of time.
(More precisely, if the Markov chain is at location $X_t$ at time $t$,  \revpoint{5}{1}
then for $x \neq y$, $\PP{X_{t+\epsilon} = y \given X_t = x} = \epsilon G_{xy} + O(\epsilon^2)$.)
For mathematical convenience, the diagonal of this matrix is completed so that rows sum to zero:
$G_{xx} = - \sum_{y \neq x} G_{xy}$.
We will need to find the \emph{hitting times} of the chain --
i.e., for each pair of states $x$ and $y$, 
the mean time until the chain first hits $y$ after being started at $x$.
We denote this quantity $H_{xy}$.
Throughout, we assume that all hitting times are finite,
which implies the chain is connected and irreducible.
The Markov property lets us write a system of linear equations that can be solved for $H$, as follows
(see, e.g., \citet{kemeny1983finite}):
Suppose that the chain begins at $x \neq z$,
and is at location $y \neq z$ after $dt$ units of time, without having visited $z$.
Then the total time to hit $z$ is equal to the elapsed time $dt$
plus the remaining time to hit $z$ from $y$.
Formally, this says that \revpoint{5}{2}
$H_{xz} \approx dt + \sum_y \P\{X_{dt} = y \given X_0 = x\} H_{yz}$.
Replacing $\P\{X_{dt} = y \given X_0 = x\}$ with $dt G_{xy}$ and rearranging
gives the system of equations:
\begin{align} \label{eqn:hitting_sum}
    \sum_y G_{xy} H_{yz} = -1 \qquad \text{for} \; x \neq z.
\end{align}
In other words, if $G_{-z}$ is the matrix with the $z^\text{th}$ row and column removed,
and $\bone$ is the vector composed of all ones,
then the $z^\text{th}$ column of $H$, except for $H_{zz}$,
can be computed as $- (G_{-z})^{-1} \bone$.
This, along with $H_{zz} = 0$, allows computation of $H$
from the movement rates $G$ \citep{kemeny1983finite}.

Suppose instead we are given the hitting times, $H$, and wish to find the movement rates, $G$.
This is not the situation we are in -- we have either commute or coalescence times --
but it is related.
The Random Target Lemma \citep{aldous-fill-2014}
tells us that the stationary distribution of the Markov chain, denoted $\pi$,
can be recovered from the hitting times by solving $\pi = H^{-1} \bone / \bone^T H^{-1} \bone$.
As shown in Appendix \ref{apx::hitting_calcs},
equation \eqref{eqn:hitting_sum} can be rewritten in matrix form as
\begin{align} \label{eqn:gh}
    G H = \diag(1/\pi) - \bone \bone^T ,
\end{align}
which implies that $G$ can be computed directly from $H$ as
\begin{align} \label{eqn:G_from_H}
    G = \left( \diag(1/\pi) - \bone \bone^T \right) H^{-1} .
\end{align}
(We prove that $H$ is invertible in Appendix~\ref{apx::hitting_calcs}.) \revpoint{1}{12}
Therefore, $H$ uniquely determines $G$.

% \paragraph{Commute times}
The \emph{commute times} are a symmetrized version of the hitting times
(which need not be symmetric),
defining $R = H + H^T$, i.e.:
\begin{align} \label{eqn:R_from_H}
\text{\bf (commute time)} \qquad
    R_{xy} = H_{xy} + H_{yx}.
\end{align}

Although commute times are uniquely determined by movement rates, the reverse is not true --
we can use equations \eqref{eqn:gh} and \eqref{eqn:R_from_H} to see how to modify movement rates
without changing commute times.  \llabel{ll:dense1}
A small concrete example of this is given in Appendix \ref{ss:simple_example}.
More generally, commute times only depend on the symmetric part of $H$ --
given a skew-symmetric matrix $Z$ (so that $Z + Z^T = 0$),
any Markov chain that has hitting times given by $H_\epsilon = H + \epsilon Z$ for some $\epsilon$
will have the same commute times.
The resulting hitting times may not be valid --
if $G_\epsilon$ is the matrix constructed by applying equation \eqref{eqn:G_from_H} to $H_\epsilon$,
% (assuming this is invertible),
then it is not guaranteed that the offdiagonal elements of $G_\epsilon$ are all nonnegative, as required.
% (It is always true that rows of $G_\epsilon$ sum to zero.)
However, since $G$ and $H$ are continuous functions of each other,
if all offdiagonal elements of $G$ are strictly positive,
then there exists a positive $\epsilon_0$ such that all $G_\epsilon$ for $\epsilon < \epsilon_0$
do define valid Markov chains, all with the same commute times.

% \paragraph{Coalescence times}
The \emph{coalescence times} are defined using \emph{two} copies of the same chain,
as the mean time until coalescence,
if the chains coalesce at rate $\gamma$ when they are in the same place.
Concretely, 
suppose that $X$ and $Y$ are independent Markov chains both moving with movement rates given by $G$,
that coalesce at rate $\gamma_x$ when $X$ and $Y$ are both at $x$.
Then we define $\tau$ to be this coalescence time,
so that 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t = Y_t = x\} = \epsilon \gamma_x + O(\epsilon^2)$,
and 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t \neq Y_t\} = O(\epsilon^2)$.
Then, the (mean) coalescence time is $C_{xy} = \E[\tau \given X_0 = x, \, Y_0 = y]$.
A similar argument using the Markov property
shows that $C$ satisfies the following equation:
\begin{align}
\text{\bf (coalescence time)} \qquad
    \sum_y \left(G_{xy} C_{yz} + C_{xy} G_{zy}\right)
    &=
    \begin{cases}
        -1                   \quad & \text{if} \; x \neq z, \\
        -1 + \gamma_z C_{zz} \quad & \text{if} \; x = z.
    \end{cases}
\end{align}
Similar recursions are common in the literature,
going back at least to \citet{hill1972effective} (see also \citet{whitlock1997effective}).
In matrix notation, this is
\begin{align} \label{eqn:C_matrix}
    G C + C G^T - \diag(\gamma) \circ C = -\bone \bone^T,
\end{align}
where $\diag(\gamma)$ is the matrix with the vector $\gamma$ on the diagonal and zeros elsewhere,
and $\circ$ is the componentwise product.
In practice, we solve this by working with the product Markov chain $(X_t, Y_t)$,
whose generator matrix is $G \otimes I + I \otimes G$, \revpoint{1}{10}
where $I$ is the identity matrix and $\otimes$ is the Kronecker product.

>>>>

The \emph{commute time} of a Markov chain between locations $i$ and $j$
is the expected time for a particle that moves according to the rules of the chain, 
started at $i$, to first reach $j$ and then return.
We will denote this by $R_{ij}$, to emphasize that it is also the effective resistance
between $i$ and $j$.
The \emph{coalescence time} is the expected time for two particles started at $i$ and $j$
to \emph{coalesce},
which happens at a given rate whenever they are in the same location.
We will denote this by $C_{ij}$.
Why should these be related to genetic distance? 
We first explain this connection and discuss computation of both quantities,
and finally develop a method that infers movement rates of the Markov chain
(which are rates of gene flow)
from observations of coalescence time. \revpoint{1}{9}
We apply the method to both realistic simulations in continuous space
and to an empirical dataset.
Although we have tried to highlight the intuitive aspects of the initial technical discussion,
readers who are interested in practical differences
more than the mathematical arguments should be able to skip to ``Test cases and simulations''
without loss of continuity. \llabel{ll:flagging}

@@

For data, we are given genotypes of individuals and their geographic coordinates;
then, we divide these individuals into groups,
and compute the genetic distance between group $i$ and group $j$
as the mean divergence between a pair of genomes, chosen one from each group.
We denote the resulting genetic distance $D_{ij}$;
since divergence is symmetric, $D_{ij} = D_{ji}$.
We compute $D_{ii}$ by sampling without replacement from group $i$,
so that generally $D_{ii} > 0$: the diagonal elements measure local diversity.
As discussed below, 
``resistance distance'' methods find commute times so that the mean commute time between
$i$ and $j$ matches $D_{ij}$ -- but the commute time of a location to itself is zero,
so a term is added to account for local diversity.
We do this in the same way as \citet{petkova2016visualizing},
by introducing the \emph{local diversity} parameters $q_i$,
and using the model
\begin{align} \label{eq:commute_approx}
	D_{ij} \approx \comdist_{ij} = R_{ij}/4 + q_{i}/2 + q_{j}/2 .
\end{align}
Concretely, this means finding 
movement rates of a Markov chain whose matrix of commute times ($R$)
yields a good approximation to the matrix of genetic distances,
after adding local diversity parameters ($q$, which are also free parameters). \revpoint{3}{7}
The factor of four is necessary to make commute time work on the same time scale as coalescence time:
the commute time covers the path from $i$ to $j$ twice
(once in both directions),
while the coalescence time covers the distance only once,
and in half the time, because two particles are moving.
% Furthermore, coalescence time includes within-location diversity
% because particles do not coalesce instantly --
% therefore, commute time is most analogous to the first meeting time.
For a conceptual picture of the differences between the coalescence time and commute time,
see Figure \ref{fig:concept_coalcom}.

@@

Now, we explain how to compute the relevant quantities --
mean commute and coalescence times -- from a Markov chain model of lineage movement.
The Markov chain is defined by its \emph{generator matrix}, denoted $G$,
for which $G_{xy}$ gives the ``jump rate'' from $x$ to $y$.
The jump rate $G_{xy}$ is \revpoint{3}{8}
the probability the chain jumps to position $y$ when at position $x$ per unit of time.
(More precisely, if the Markov chain is at location $X_t$ at time $t$,  \revpoint{5}{1}
then for $x \neq y$, $\PP{X_{t+\epsilon} = y \given X_t = x} = \epsilon G_{xy} + O(\epsilon^2)$.)
For mathematical convenience, the diagonal of this matrix is completed so that rows sum to zero:
$G_{xx} = - \sum_{y \neq x} G_{xy}$.
We will need to find the \emph{hitting times} of the chain --
i.e., for each pair of states $x$ and $y$, 
the mean time until the chain first hits $y$ after being started at $x$.
We denote this quantity $H_{xy}$.
Throughout, we assume that all hitting times are finite,
which implies the chain is connected and irreducible.
The Markov property lets us write a system of linear equations that can be solved for $H$, as follows
(see, e.g., \citet{kemeny1983finite}):
Suppose that the chain begins at $x \neq z$,
and is at location $y \neq z$ after $dt$ units of time, without having visited $z$.
Then the total time to hit $z$ is equal to the elapsed time $dt$
plus the remaining time to hit $z$ from $y$.
Taking the average over possible paths, this says that \revpoint{5}{2}
$H_{xz} \approx dt + \sum_y \P\{X_{dt} = y \given X_0 = x\} H_{yz}$.
Replacing $\P\{X_{dt} = y \given X_0 = x\}$ with $dt G_{xy}$ and rearranging
gives the system of equations:
\begin{align} \label{eqn:hitting_sum}
    \sum_y G_{xy} H_{yz} = -1 \qquad \text{for} \; x \neq z.
\end{align}
In other words, if $G_{-z}$ is the matrix with the $z^\text{th}$ row and column removed,
and $\bone$ is the vector composed of all ones,
then the $z^\text{th}$ column of $H$, except for $H_{zz}$,
can be computed as $- (G_{-z})^{-1} \bone$.
This, along with $H_{zz} = 0$, allows computation of $H$
from the movement rates $G$ \citep{kemeny1983finite}.

Suppose instead we are given the hitting times, $H$, and wish to find the movement rates, $G$.
This is not the situation we are in -- we have either commute or coalescence times --
but it is related.
The Random Target Lemma \citep{aldous-fill-2014}
tells us that the stationary distribution of the Markov chain, denoted $\pi$,
can be recovered from the hitting times by solving $\pi = H^{-1} \bone / \bone^T H^{-1} \bone$.
As shown in Appendix \ref{apx::hitting_calcs},
equation \eqref{eqn:hitting_sum} can be rewritten in matrix form as
\begin{align} \label{eqn:gh}
    G H = \diag(1/\pi) - \bone \bone^T ,
\end{align}
which implies that $G$ can be computed directly from $H$ as
\begin{align} \label{eqn:G_from_H}
    G = \left( \diag(1/\pi) - \bone \bone^T \right) H^{-1} .
\end{align}
(We prove that $H$ is invertible in Appendix~\ref{apx::hitting_calcs}.) \revpoint{1}{12}
Therefore, the pairwise mean hitting times, $H$, uniquely determine the jump rates, $G$.

% \paragraph{Commute times}
The \emph{commute time} between $x$ and $y$ is the mean time a chain started at $x$ takes 
to first hit $y$ and then to return to $x$.
We will write this as
\begin{align} \label{eqn:R_from_H}
\text{\bf (commute time)} \qquad
    R_{xy} = H_{xy} + H_{yx},
\end{align}
or in matrix notation, $R = H + H^T$.
Note that the commute times are symmetric,
even though the hitting times may not be.

Although commute times are uniquely determined by movement rates, the reverse is not true --
we can use equations \eqref{eqn:gh} and \eqref{eqn:R_from_H} to see how to modify movement rates
without changing commute times.  \llabel{ll:dense1}
A small concrete example of this is given in Appendix \ref{ss:simple_example}.
More generally, commute times only depend on the symmetric part of $H$ --
given a skew-symmetric matrix $Z$ (so that $Z + Z^T = 0$),
any Markov chain that has hitting times given by $H_\epsilon = H + \epsilon Z$ for some number $\epsilon$
will have the same commute times.
The resulting hitting times may not be valid --
if $G_\epsilon$ is the matrix constructed by applying equation \eqref{eqn:G_from_H} to $H_\epsilon$,
% (assuming this is invertible),
then it is not guaranteed that the offdiagonal elements of $G_\epsilon$ are all nonnegative, as required.
% (It is always true that rows of $G_\epsilon$ sum to zero.)
However, since $G$ and $H$ are continuous functions of each other,
if all offdiagonal elements of $G$ are strictly positive,
then there exists a positive $\epsilon_0$ such that all $G_\epsilon$ for $\epsilon < \epsilon_0$
do define valid Markov chains, all with the same commute times.
